# Missing data and Imputation

Imputation is the process of estimating missing data points. This can be done in any number of ways, and as usual, the "best" way depends on the problem.

Of course, you don't *have* to impute data. You can also simply delete any indicator or unit that has missing values, although in many cases this can be too restrictive. Reasonable results can still be obtained despite small amounts of missing data, although if too much data is missing, the uncertainty can be too high to give a meaningful analysis. As usual, it is a balance.

A good first step is to check how much data is missing, and where (see below). Units with very high amounts of missing data can be screened out. Small amounts of missing data can then be imputed.

## Concept

The simplest imputation approach is to use values of other units to estimate the missing point. Typically, this could involve the sample mean or median of the indicator. Here's some data from the ASEM data set regarding the average connection speed of each country. Towards the end there are some missing values, so let's view the last few rows:

```{r}
library(COINr)
Ind1 <- data.frame(Country = ASEMIndData$UnitName, ConSpeed = ASEMIndData$ConSpeed)
Ind1[40:nrow(Ind1),]
```

Using our simple imputation method, we just replace the `NA` values with the sample mean.

```{r}
Ind1$ConSpeed <- replace(Ind1$ConSpeed, is.na(Ind1$ConSpeed), mean(Ind1$ConSpeed, na.rm = T))
Ind1[40:nrow(Ind1),]
```

This approach can be reasonable if countries are somehow similar in that indicator. However, other perhaps more informative ways are available:

* Substituting by the mean or median of the country group (e.g. income group or continent)
* If time series data is available, use the latest known data point

Then there is another class of methods which use data from other indicators to estimate the missing point. The core idea here is that if the indicators are related to one another, it is possible to guess the missing point by using known values of other indicators. This can take the form of:

* Simply substituting the mean or median of the *normalised* values of the other indicators
* Subsituting the mean or median of *normalised* values of the other indicators *within the aggregation group*
* Using a more formal approach, based on regression or more generally on statistical modelling

Let's explore the options available in COINr

## Data checks and screening

A first step is to check in detail how much data missing. The function `checkData()` does this, and also has the option to screen units based on data availability.

```{r, collapse=T}
# Assemble ASEM data first
ASEM <- assemble(IndData = COINr::ASEMIndData, IndMeta = COINr::ASEMIndMeta,
                 AggMeta = COINr::ASEMAggMeta)
# Missing data check on the raw data set
ASEM <- checkData(ASEM, dset = "Raw")
head(ASEM$Analysis$Raw$MissDatSummary)
```

*Note: to include updated checkData() function with screening for zeros.*

The `MissDataSummary` table shows the unit code, number of missing observations, and overall percentage data availability. Finally, the `LowDataAll` column can be used as a flag for units with data availability lower than a set amount `ind_thresh` which is one of the input arguments to `checkData()`. ASEM indicators were already chosen to fulfill minimum data requirements, for which reason the data availability is all above the default of 2/3.

We will set the minimum data threshold a bit higher, and in doing so also demonstrate another feature of `checkData()`, which is to automatically screen units based on data availability. Setting `unit_screen = TRUE` will generate a new data set `.$Data$Screened` which only includes units with data availability above the set threshold. This data set can then be passed on to subsequent operations.

```{r}
# Missing data check on the raw data set
ASEM <- checkData(ASEM, dset = "Raw", ind_thresh = 0.85, unit_screen = TRUE)
ASEM$Analysis$Raw$RemovedUnits
```

This generates a new data set `.$Data$Screened` with the removed units recorded in `.$Analysis$Raw$RemovedUnits`, which in this case are Brunei, Laos and Myanmar. As a final option to mention, you can manually include or exclude units. So for example, a unit that doesn't have sufficient data coverage could be manually included anyway, and another unit could be excluded.

```{r}
# Countries to include and exclude
df <- data.frame(UnitCode = c("AUT","BRN"), Status = c("Exclude", "Include"))

# Missing data check on the raw data set
ASEM <- checkData(ASEM, dset = "Raw", ind_thresh = 0.85, unit_screen = TRUE, Force = df)
ASEM$Analysis$Raw$RemovedUnits
```

Now AUT has been excluded, even though it had sufficient data coverage, and BRN has been manually included, even though it didn't. The intention is to give some flexibility to make exceptions for hard rules.

The other output of interest from `checkData()` is missing data by group:

```{r}
head(ASEM$Analysis$Raw$MissDatByGroup[30:40,])
```

This gives the percentage indicator data availability inside each aggregation group.

## Imputation in COINr

Now we turn to estimating missing data points in COINr. The function of interest is `impute()`, which gives a number of options, corresponding to the types of imputation discussed above. Briefly, COINr can impute using:

* The indicator mean, either across all countries or within a specified group
* The indicator median, either across all countries or within a specified group
* The mean or median, within the aggregation group
* The expectation maximisation algorithm, via the [AMELIA](https://cran.r-project.org/web/packages/Amelia/index.html) package

### By column

The `impute()` function follows a similar logic to other COINr functions. We can enter a COIN or a data frame, and get a COIN or a data frame back. Let's impute on a COIN first, using the raw ASEM data.

```{r, collapse=T}
# Assemble ASEM data if not done already (uncomment the following)
# ASEM <- assemble(IndData = COINr::ASEMIndData, IndMeta = COINr::ASEMIndMeta,
#                 AggMeta = COINr::ASEMAggMeta)

# check how many NAs we have
sum(is.na(ASEM$Data$Raw))

# impute using indicator mean
ASEM <- impute(ASEM, dset = "Raw", imtype = "ind_mean")

# check how many NAs after imputation
sum(is.na(ASEM$Data$Imputed))
```

If the input is a COIN, by default the output is an updated COIN, with a new data set `.$Data$Imputed`. We can also set `out2 = "df"` to output the imputed data frame on its own.

The above example replaces `NA` values with the mean of the indicator, and setting `imtype = ind_median` will instead use the indicator median. In this respect, we are using information from other units, inside the same indicator, to estimate the missing values.

Imputing by group may be more informative if you have meaningful grouping variables. For example, the ASEM data set has groupings by GDP, population, GDP per capita and whether the country is European or Asian. We might hypothesise that it is better to replace `NA` values with the median inside a group, say by GDP group. This would imply that countries with a similar GDP are likely to have similar indicator values.

```{r, collapse=T}
# impute using GDP group median
ASEM <- impute(ASEM, dset = "Raw", imtype = "indgroup_median", groupvar = "Group_GDP")
```

### By row

So far, the imputation has been using values from the same column (indicator). Another possibility is to use values from other indicators, i.e operate row-wise. COINr offers two basic options in this respect: either to take the mean or median of the other indicators inside the aggregation group, and this can be done by setting `imtype = "agg_mean"` or `imtype = "agg_median"` respectively.

Both of these imputation methods will, for a given unit, take the mean or median of the *normalised* values of the other indicators in the aggregation group in the level above indicator level. The process is as follows, for each `NA` value:

1. Find indicators in the same aggregation group
2. Normalise these indicators using the min-max method, so they each have a minimum of zero and a maximum of one.
3. Replace the `NA` with the mean or median of the normalised values, for the unit in question.
4. Reverse the min-max transformation so that the indicators are returned to their original scale.

It's important to realise that this is equivalent to aggregating with the mean or median without imputing first. This is because in the aggregation step, if we take the mean of a group of indicators, and there is a `NA` present, this value is excluded from the mean calculation. Doing this is mathematically equivalent to assigning the mean to that missing value. Therefore, one reason to use this imputation method is to see which values are being implicitly assigned as a result of excluding missing values from the aggregation step. This is sometimes known as "shadow imputation".

### By column and row

Finally, we can use a regression-based *expectation maximisation* algorithm, imported via the `AMELIA` package. This estimates missing values using the values of the other indicators *and* the other units. This effectively involves regressing each indicator on other indicators, and predicting missing values.

A catch of this approach is that if the number of units is small compared to the number of indicators, there will not be enough observations to estimate the parameters of the model. This is definitely the case if you have more indicators than observations. To use the EM algorithm in this case, imputation is performed on aggregation groups. So, to use the EM approach in `COINr`, you have to specify the aggregation level to group indicators by.

```{r, collapse=T}
# impute using EM, grouping indicators inside their pillars (level 2)
ASEM <- impute(ASEM, imtype = "EM", EMaglev = 2)
```

This works, because the ASEM data set has 51 units and no more thann eight indicators per pillar. However, if we were to use `EMaglev = 3`, i.e. try imputing indicators grouped into the two sub-indexes, it does not work.

Finding the right aggregation level to impute at might involve a bit of trial and error, but the advantage of imputing by groups of indicators is that aggregation groups are usually composed of indicators that are similar in some respect. This makes it more likely that they are good predictors of the other indicators in their group.

The `AMELIA` package offers many more options for imputation that are not available through the `impute()` function, such as bootstrapping, time-series imputation and more. As with plotting packages `COINr` aims to provide an easy and quick interface to standard imputation in `AMELIA`, but if you want to have full control you should use the `AMELIA` package directly. Other good imputation options are the [`MICE`](https://cran.r-project.org/web/packages/mice/index.html) package (Multivariate Imputation via Chained Equations) and the [`missForest` ](https://cran.rstudio.com/web/packages/missForest/index.html) package (using random forests).
