# Helper functions

## R Interfaces

### Data import

COINr has a couple of useful functions to help import and export data and metadata. You might have heard of the [COIN Tool](https://knowledge4policy.ec.europa.eu/composite-indicators/coin-tool_en) which is an Excel-based tool for building and analysing composite indicators, similar in fact to COINr^[Full disclosure, I was also involved in the development of the COIN Tool]. With the `coinToolIn()` function you can import data directly from the COIN Tool to cross check or extend your analysis in COINr.

To demonstrate, we can take the example version of the COIN Tool, which you can download [here](https://composite-indicators.jrc.ec.europa.eu/sites/default/files/COIN_Tool_v1_LITE_exampledata.xlsm). Then it's as simple as running:

```{r Import CT}
library(COINr6)

# This is the file path and name where the COIN Tool is downloaded to
# You could also just put it in your project directory.
fname <- "C:/Users/becke/Downloads/COIN_Tool_v1_LITE_exampledata.xlsm"

dflist <- COINToolIn(fname)
```

The output of this function is a list with the three data frame inputs to `assemble()`.

```{r}
dflist
```

Because the COIN Tool uses numeric codes for indicators such as `ind.01`, you might want slightly more informative codes. The best way to do this is to name the codes yourself, but a quick solution is to set `makecodes = TRUE` in `COINToolIn()`. This generates short codes based on the indicator names. It will not yield perfect results, but for a quick analysis it might be sufficient. At least, you could use this and then modify the results by hand.

```{r}
dflist <- COINToolIn(fname, makecodes = TRUE)
dflist$IndMeta$IndCode
```

While the codes could certainly be improved, it's a lot better than uninformative numbers. Finally, we can assemble the output into a COIN and begin the construction.

```{r, collapse=T}
ESI <- assemble(dflist[[1]],dflist[[2]],dflist[[3]],)
```

### Export to Excel

Trigger warning for R purists! Sometimes it's easier to look at your data in Excel. There, I said it. R is great for doing all kinds of complicated tasks, but if you just want to look at big tables of numbers and play around with them, maybe make a few quick graphs, then Excel is a great tool.

Actually Excel is kind of underrated by many people who are used to programming in R or Python, Matlab or even Stata. It has a lot of clever tools that not many people know about. But more importantly, Excel is a *lingua franca* between all kinds of professions - you can pass an Excel spreadsheet to almost anyone and they will be able to take a look at it and use the data. Try doing that with an R or Python script.

It just boils down to using the right tool for the right job. Anyway, with that aside, let's look at COINr's `coin2Excel()` function. You put in your COIN, and it will write a spreadsheet.

```{r, eval=F}
# Build ASEM index
ASEM <- build_ASEM()
# Get some statistics
ASEM <- getStats(ASEM, dset = "Raw")
# write to Excel
coin2Excel(ASEM, fname = "ASEMresults.xlsx")
```

The spreadsheet will contain a number of tabs, including:

* The indicator data, metadata and aggregation metadata that was input to COINr
* All data sets in the `.$Data` folder, e.g. raw, treated, normalised, aggregated, etc.
* (almost) All data frames found in the `.$Analysis` folder, i.e. statistics tables, outlier flags, correlation tables.

## Selecting data sets and indicators

The `getIn()` function is widely used by many `COINr` functions. It is used for selecting specific data sets, and returning subsets of indicators. While some of this can be achieved fairly easily with base R, or `dplyr::select()`, subsetting in a hierarchical context can be more awkward. That's where `getIn()` steps in to help.

Although it was made to be used internally, it might also help in other contexts. Note that this can work on COINs or data frames, but is most useful with COINs.

Let's take some examples. First, we can get a whole data set. `getIn()` will retrieve any of the data sets in the `.$Data` folder, as well as the denominators.

```{r}
# Build data set first, if not already done
ASEM <- build_ASEM()

# Get raw data set
datalist <- getIn(ASEM, dset = "Raw")
datalist$ind_data_only
```

The output, here `datalist` is a list containing the full data set `.$ind_data`, the data set `.$ind_data_only` only including numerical (indicator) columns, as well as unit codes, indicator codes and names, and the object type.

More usefully, we can get specific indicators:

```{r}
# Get raw data set
datalist <- getIn(ASEM, dset = "Raw", icodes = c("Flights", "LPI"))
datalist$ind_data_only
```
More usefully still, we can get groups of indicators based on their groupings. For example, we can ask for indicators that belong to the "Political" group:

```{r}
# Get raw data set
datalist <- getIn(ASEM, dset = "Raw", icodes = "Political", aglev = 1)
datalist$ind_data_only
```

To do this, you have to specify the `aglev` argument, which specifies which level to retrieve the indicators from. Before the data set is aggregated, you can anyway only select the indicators, but for the aggregated data set, the situation is more complex. To illustrate, we can call the Connectivity sub-index, first asking for all indicators:

```{r}
# Get raw data set
datalist <- getIn(ASEM, dset = "Aggregated", icodes = "Conn", aglev = 1)
datalist$ind_data_only
```
Or we can call all the pillars belonging to "Connectivity", i.e. the level below:

```{r}
# Get raw data set
datalist <- getIn(ASEM, dset = "Aggregated", icodes = "Conn", aglev = 2)
datalist$ind_data_only
```

Finally, if we want "Conn" itself, we can just call it directly with no `aglev` specified.

```{r}
# Get raw data set
datalist <- getIn(ASEM, dset = "Aggregated", icodes = "Conn")
datalist$ind_data_only
```

We can also use `getIn()` with data frames, and it will behave in more or less the same way, except a data frame has no information about the structure of the index. Here, `getIn()` returns what it can, and arguments like `dset` and `aglev` are ignored.

```{r}
# use the ASEM indicator data frame directly
datalist <- getIn(ASEMIndData, icodes = c("LPI", "Goods"))
datalist$ind_data_only
```

## Data frame tools

The `roundDF()` function is a small helper function for rounding data frames that contain a mix of numeric and non-numeric columns. This is very handy for presenting tables generated by COINr in documents.

```{r}
# use the ASEM indicator data frame directly
ASEMIndData |>
  roundDF(decimals = 3) |>
  reactable::reactable(defaultPageSize = 5, highlight = TRUE, wrap = F)
```

By default, numbers are rounded to two decimal places.

A similar function, `rankDF()`, converts a data frame of numbers/scores to ranks, ignoring any non-numeric columns.

```{r}
ASEMIndData[c("UnitCode", "LPI", "Goods")] |>
  rankDF() |>
  head(10)
```

Note that there are different ways to rank values, depending on how you deal with ties. The `rankDF()` function uses so-called "sport" ranking, where ties are assigned the highest rank value. See `?rank` for some information on this.

In-group ranking is also possible using `rankDF()` by specifying a column of the data frame to use as a grouping variable.

```{r}
ASEMIndData[c("UnitCode", "Group_GDP", "Goods", "LPI")] |>
  rankDF(use_group = "Group_GDP") |>
  head(10)
```

A final helpful function is `compareDF()`, which gives a detailed comparison of two similar data frames. This is particularly useful when comparing results from one calculation to another. For example, let's say you have built a composite indicator in Excel, but want to replicate it in COINr, to use some of the COINr tools. But having built it, you want to be sure that the results are exactly the same. From experience, it is actually quite common for differences to appear, which could be due to calculation errors, but also sometimes due to differences in the way that Excel calculates certain things. Anyway, how can you check that the two sets of results are the same, and if not, what are the differences?

The `compareDF()` function is intended to compare two data frames that have the same variables and rows, but possibly in a different order. Moreover it allows to compare at a specified number of significant figures and will tell you specifically what the differences are. This is probably best clarified with an example.

```{r}
# take a sample of indicator data (including the UnitCode column)
data1 <- ASEMIndData[c(2,12:15)]
# copy the data
data2 <- data1
# make a change: replace one value in data2 by NA
data2[1,2] <- NA
# compare data frames
compareDF(data1, data2, matchcol = "UnitCode")
```

The `compareDF()` function requires a column to be specified which is used to match rows. It will then check:

* Whether the two data frames are the same size
* Whether the column names are the same, and that the matching column has the same entries
* Column by column that the elements are the same, after sorting according to the matching column

The output summarises whether the two data frames are the same (`TRUE`/`FALSE`), which columns are different and how, and finally identifies specific data points which are different and returns both values.
