# Appendix: Analysing a Composite Indicator Example

Here some possible steps for analysing a composite indicator are given. The format for doing this might depend on what you are trying to accomplish. If you want to analyse your own composite indicator as you are building it, this should be integrated with the previous [Appendix: Building a Composite Indicator Example]. If you are analysing an existing composite indicator built by someone else, you may wish to do this in a separate R Markdown document, for example.

## Loading data

If the analysis is part of constructing a composite indicator in COINr, you should have a COIN assembled following the steps in the previous chapter. However, if you are analysing an existing composite indicator, there may be a slight difference. It is possible, for example, that you may be provided with indicator data, which may already be normalised and aggregated.

COINr's `assemble()` function allows you to assemble a COIN with pre-aggregated data. This is done by setting `preagg = TRUE`. If this is enabled, COINr will create a COIN with a data set called "PreAggregated". To do this, you need a pre-aggregated data set which includes (possibly normalised) columns for all indicators *and* for all aggregates.

To demonstrate, we can use a pre-aggregated data set created by COINr.

```{r, message=FALSE}
library(COINr)

# build ASEM index
ASEM <- build_ASEM()

# extract aggregated data set (as a data frame)
Aggregated_Data <- ASEM$Data$Aggregated

# assemble new COIN only including pre-aggregated data
ASEM_preagg <- assemble(IndData = Aggregated_Data,
                        IndMeta = ASEMIndMeta,
                        AggMeta = ASEMAggMeta,
                        preagg = TRUE)
```

COINr will check that the column names in the indicator data correspond to the codes supplied in the `IndMeta` and `AggMeta`. This means that these two latter data frames still need to be supplied. However, from this point the COIN functions as any other, although consider that it cannot be regenerated (the methodology to arrive at the pre-aggregated data is unknown), and the only data set present is the "PreAggregated" data.

## Check calculations

If you are using pre-aggregated data, you may wish to check the calculations to make sure that they are correct. If you additionally have raw data, and you know the methodology used to build the index, you can recreate this by rebuilding the index in COINr. If you only have normalised and aggregated data, you can still at least check the aggregation stage as follows. Assuming that the indicator columns in your pre-aggregated data are normalised, we can first manually create a normalised data set:

```{r}
library(dplyr)

ASEM_preagg$Data$Normalised <- ASEM_preagg$Data$PreAggregated %>%
  select(!ASEM$Input$AggMeta$Code)
```

Here we have just copied the pre-aggregated data, but removed any aggregation columns.

Next, we can aggregate these columns using COINr.

```{r}
ASEM_preagg <- aggregate(ASEM_preagg, dset = "Normalised", agtype = "arith_mean")

# check data set names
names(ASEM_preagg$Data)
```

Finally, we can check to see whether these data frames are the same. There are many possible ways of doing this, but a simple way is to use dplyr's `all_equal()` function.

```{r}
all_equal(ASEM_preagg$Data$PreAggregated,
          ASEM_preagg$Data$Aggregated)
```

As expected, here the results are the same. If the results are *not* the same, `all_equal()` will give some information about the differences. If you reconstruct the index from raw data, and you find differences, a few points are worth considering:

1. The difference could be due to an error in the pre-aggregated data, or even a bug in COINr. If you suspect the latter please open an issue on the repo.
2. If you have used data treatment or imputation, differences can easily arise. One reason is that some things are possible to calculate in different ways. COINr uses certain choices, but other choices are also valid. Examples of this include:
    - Skew and kurtosis (underlying data treatment) - see e.g. `?e1071::skewness`
    - Correlation and treatment of missing values - see `?cor`
    - Ranks and how to handle ties - see `?rank`
3. Errors can also arise from how you entered the data. Worth re-checking all that as well.

Double-checking calculations is tedious but in the process you often learn a lot.

## Indicator statistics

One next step is to check the indicator statistics. If you have the raw data, it is probably advisable to do this both on the raw data, and the normalised/aggregated data. Here, we will just do this on the aggregated data (but this works on any data set, e.g. also the "PreAggregated" and "Normalised" data sets created above):

```{r}
ASEM <- getStats(ASEM, dset = "Aggregated")

# view stats table
# this is rounded first (to make it easier to view), then sent to reactable to make an interactive table
# you can also view this in R Studio or use any other table viewer to look at it.
ASEM$Analysis$Aggregated$StatTable %>%
  roundDF() %>%
  reactable::reactable()
```

Particular things of interest here will be whether indicators or aggregates are highly skewed, the percentage of missing data for each indicator, the percentage of unique values and zeroes, the presence of correlations with denominators, and negative correlations. Recall that `getStats()` allows you to change thresholds for flagging outliers and high/low correlations.

Another output of `getStats()` is correlation matrices, which are also found in `ASEM$Analysis$Aggregated`. A good reason to include these here is that as part of the COIN, we can export all analysis to Excel, if needed (see the end of this chapter). However, for viewing correlations directly, COINr has dedicated functions - see the next section.

Before arriving at correlations, let's also check data availability in more detail. The function of interest here is  `checkData() `.

```{r}
ASEM <- checkData(ASEM, dset = "Raw")

# view missing data by group
ASEM$Analysis$Raw$MissDatByGroup %>%
  roundDF %>%
  reactable::reactable()
```

This adds analysis tables to the `ASEM$Analysis` folder (analysis always appears under the respective "dset" name). We have applied the analysis to the raw data set because in later steps, the data is imputed. Here, the missing data is given by each aggregation group. This helps to flag, for instance, cases where there is very low data availability for a unit in a given aggregation group. We can check the minimum:

```{r}
# get minimum, exclude first column (not numerical)
min(ASEM$Analysis$Raw$MissDatByGroup[-1])
```

This shows that in the worst case, there is more than 50% data availability for each unit in each aggregation group.

## Multivariate analysis

This section will overlap to some extent with the [Multivariate analysis] chapter. In any case we will summarise a few options here at the risk of repetition.

Perhaps the most useful way to view correlations for an aggregated index is to look at correlations of each indicator or aggregate with its parents. This is done quickly in COINr with the `plotCorr()` function.

```{r}
plotCorr(ASEM, dset = "Aggregated", showvals = T, withparent = "family", flagcolours = TRUE)
```

This shows at a glance where some problems may lie. In particular we learn that the TBT indicator is negatively correlated with all its parent groups. We can also see that the "Forest" indicator is insignificantly correlated with its parents. This table is generated here for indicators, but can also generated for other levels by setting the `aglevs` argument.

```{r}
plotCorr(ASEM, dset = "Aggregated", showvals = T, withparent = "family", flagcolours = TRUE, aglevs = 2)
```

Recall also that this function can return data frames for your own processing or presentation, rather than figures, by changing the `out2` argument.

You may also want to view correlations between indicators or within specific groups. The `plotCorr()` function is flexible in this respect.

```{r}
# plot correlations within a specific pillar (Physical)
plotCorr(ASEM, dset = "Raw", icodes = "Physical", aglevs = 1, cortype = "spearman", pval = 0)
```

Next, let's


*To be completed*
